Longitudinal Language Analysis in the Diagnosis of Mild Cognitive Impairment: A Case Study

### 2019-02-07 Source/Presidents Corpus Paper/paper1.tex
Paper1 update

\author{
  Jomar C. Alcantara \\
  Department of Computer Science \\
  School of Engineering and Applied Sciences\\
  Aston University\\
  Birmingham, UK \\
  \texttt{alcantaj@aston.ac.uk} \\
  \And
  Peter Sawyer \\
  Department of Computer Science \\
  School of Engineering and Applied Sciences \\
  Aston University\\
  Birmingham, UK \\
  \texttt{p.sawyer@aston.ac.uk} \\
  \And
  Cristina Romani \\
  Department of Psychology \\
  School of Life and Health Science\\
  Aston University\\
  Birmingham, UK \\
  \texttt{c.romani@aston.ac.uk} \\
}
% HEADER ENDS
% DOCUMENT BEGINS
\begin{document}
\maketitle
\section{Abstract}

@@

In 2009, the Department of Health designed it's National Dementia strategy and as part of this made early diagnosis and support one of it's key priorities \cite{England2009}. A lot of work has gone into trying to find ways of improving the early diagnosis of Alzheimer's Disease (AD) and Mild Cognitive Impairment (MCI) with research focused on two areas - identifying biological markers and analyzing the cognitive decline of those who are suspected to have the disease \cite{Taler2008}. As described above, the numbers of those suffering from AD and MCI are going to increase as the population ages \cite{Prince2015} and thus it is important that we utilize technology wherever possible to aid clinicians in the detection of MCI and AD. At the present time diagnosis is typically conducted at memory clinics by trained clinicians \cite{Boschi2017}. I theorize that we may be able to enable an earlier diagnosis of those with MCI and AD using samples of spontaneous speech, natural language processing (NLP) and machine learning (ML).

@@

There is a large body of research that looks at the decline in language of those with MCI and AD \cite{Taler2008, Boschi2017}. However there is conflicting evidence in these studies about which declining language factors are associated with MCI and AD \cite{Taler2008, Boschi2017}. Research therefore should look at these features in more detail and a clarification of this currently disorganised picture should go some way to helping researchers further understand the disease and it's progression. There is some evidence to support this, Berisha et al \cite{Berisha2015}, has shown through a longitudinal language analysis of spontaneous speech that there are marked differences in different parts of speech in this process between those who would go on to have a diagnosis of AD and a healthy control. 

@@

\begin{thebibliography}{10}

@@

@@

\bibitem{Wendelstein2014}
Britta Wendelstein, Ekkehard Felder, J{\"{o}}rn Stegmeier, and Johannes
  Schr{\"{o}}der.
\newblock {Noun or Pronoun?: Using Parts of Speech in ALzheimer's Disease}.
\newblock {\em Alzheimer's {\&} Dementia}, 10(4):P434, jul 2014.

>>>>

\author{
  Jomar C. Alcantara \\
  Department of Computer Science \\
  School of Engineering and Applied Sciences\\
  Aston University\\
  Birmingham, UK \\
  \texttt{alcantaj@aston.ac.uk} \\
  \And
  Peter Sawyer \\
  Department of Computer Science \\
  School of Engineering and Applied Sciences \\
  Aston University\\
  Birmingham, UK \\
  \texttt{p.sawyer@aston.ac.uk} \\
  \And
  Cristina Romani \\
  Department of Psychology \\
  School of Life and Health Science\\
  Aston University\\
  Birmingham, UK \\
  \texttt{c.romani@aston.ac.uk} \\
}
% HEADER ENDS
% DOCUMENT BEGINS
\begin{document}
\maketitle
\begin{abstract}
Dementia is a neurological disorder that has been consistently growing in numbers over time. The analysis of language has been highlighted as a potential method for diagnosing Alzheimer's disease at its early or pre-clinical stages. This paper builds on the work of Berisha and Liss who looked at extracting language features from the non-scripted portions of press conferences from two US presidents, President Ronald Reagan, who would later go on to receive a diagnosis of Alzheimer's Disease and President George Herbert Walker Bush who received no such diagnosis. In this paper, we add the transcripts of President Donald John Trump who in terms of age is a much more appropriate match and we extract a larger variety of language features which have been shown to be associated with language decline due to Dementia. Once the features were extracted we completed a regression analyses. We found that President Reagan showed a significant decrease in unique words over time and his usage of nouns and adjectives. We also found that there was an increase in his usage of non-specific nouns, adverbs and pronouns. We found no significant trends in the transcripts of Presidents Bush and Trump. This study shows that language deterioration can be tracked over time in those with Mild Cognitive Impairment and is generally differentiable from healthy controls. The results also potentially demonstrate the utility of longitudinal language analysis as a method of diganosing Mild Cognitive Impairment in the future.
\end{abstract}
\keywords{Mild Cognitive Impairment \and Alzheimer's Disease \and Natural Language Processing \and Longitudinal \and Early Diagnosis \and Language}

@@

In 2009, the UK's Department of Health designed it's National Dementia strategy and as part of this made early diagnosis and support one of it's key priorities \cite{England2009}. A lot of work has gone into this with research focused on two areas - identifying biological markers and analyzing the cognitive decline of those who are suspected to have the disease \cite{Taler2008}. Recently, researchers have highlighted that technology, more specifically the domains of Natural Language Processing (NLP) and Machine Learning (ML), is ideally placed to aid in this endeavour \cite{Boschi2017}. It is theorized that we may be able to use these technologies to enable an earlier diagnosis of those with MCI and AD using samples of spontaneous speech, natural language processing (NLP) and machine learning (ML) \cite{Orimaye2014, Boschi2017}.

@@

There is a large body of research that looks at the decline in language of those with MCI and AD \cite{Taler2008, Boschi2017, Emery2000}. However there is conflicting evidence in these studies about which declining language factors are associated with MCI and AD \cite{Taler2008, Boschi2017}. Some work therefore should look at these features in more detail and a clarification of this currently disorganised picture should go some way to helping researchers further understand the disease and it's progression. There is some evidence to support this, Berisha et al \cite{Berisha2015}, has shown through a longitudinal language analysis of spontaneous speech that there are marked differences in different parts of speech in this process between those who would go on to have a diagnosis of AD and a healthy control. 

@@

\bibliographystyle{unsrt}
\begin{thebibliography}{10}

@@

\bibitem{Emery2000}
Olga~B. Emery.
\newblock {Language Impairment in Dementia of the Alzheimer's Type: A
  Hierarchical Decline?}
\newblock {\em International Journal of Psychiatry in Medicine},
  30(2):145--164, jul 2000.

@@

\bibitem{Wendelstein2014}
Britta Wendelstein, Ekkehard Felder, J{\"{o}}rn Stegmeier, and Johannes
  Schr{\"{o}}der.
\newblock {NOUN OR PRONOUN?: USING PARTS OF SPEECH IN ALZHEIMER'S DISEASE}.
\newblock {\em Alzheimer's {\&} Dementia}, 10(4):P434, jul 2014.
